Donor Preferences in Canadian Political Financing

### 2024-12-14 paper/paper.qmd
update discussion and references

A key question in political finance is whether a party's position in
power affects the amount of financial support it receives. Studies in
other countries, such as the United States, have shown that incumbents
often benefit from more donations, particularly from interest groups
seeking to maintain influence (@fouirnaies2014financial). However, there
has been less focus on whether this holds true in Canada, where the
political system and donation regulations differ significantly. This
paper aims to fill that gap by analyzing how a party's power status
influences donor behaviour in federal and Ontario provincial elections.

@@

The analysis was conducted using both R (@citeR) and Python (@python),
leveraging a range of libraries for data manipulation and visualization.
In R, key packages included dplyr (@dplyr), lubridate (@lubridate),
tidyverse (@tidyverse), ggplot2 (@ggplot), readr (@readr), tibble
(@tibble), and rstanarm (@rstanarm) for statistical modeling. In Python,
pandas (@pandas), requests (@requests), and matplotlib (@matplotlib)
were utilized for gathering the data.

@@

At the heart of politics is finance, as revealed by Traag's paper on the
complex contagion of campaign donations @traag2016complex. Their study
demonstrates that a majority of campaign contributions in the US
originate from a few wealthy individuals through large social networks.
With the constant battle between government and corporations, the
findings in this study underscore the significant influence of financial
power in shaping political dynamics.

The landscape in Canada for political financing is growing each year.
Results from studies demonstrate that individual contributors may prefer
to support challengers for several reasons. An optimistic but naive
explanation may be that individuals hope to foster competitive
elections, given that donations depend not on political leaning but on
who is currently in power. However, a more realistic explanation for
this phenomenon is that some individuals want to make policy changes. It
is no secret that politicians, particularly those out of power, often
solicit funding and political support in exchange for advancing the
interests of their stakeholders @mollona2022ethics. These two dynamics
may be working in tandem to create an effect where parties eager to be
elected rely more on funding from external sources, which can switch
each election cycle.

Moreover, these trends of political donations fluctuating based on the
party in power are not unique to Canada. The study "Financing UK
Democracy: A Stocktake of 20 Years of Political Donations" highlights a
surge in private political donations, more specifically from
superdonors, widening the resource gap between large parties
@draca2022financing. Similar to Canada, the UK also sees large
donations from a wealthy few influencing their elections each year. The
parallel between Canada and the UK underscores a broader democratic
issue of financial power being used to influence political outcomes.
Though this concept is not novel, this study contributes to the already
growing spotlight on the persistent lack of proper reform in political
systems.

@@

This study provides valuable insights into donation behaviours across
Canada but it its limitations. First, the analysis is applied to only
federal and provincial donation records, overlooking regional variations
found in other provinces and territories. The geographical limitations
of this study hinder its ability to generalize across the entire
country. Moreover, the study only spans from 2006 to 2024 which economic
or political events could have influence over the donation patterns that
the models do not account for. Additionally qualitative factors such as
personal beliefs, specific policy preferences and campaign effectiveness
are not directly addressed in this study. Data quality issues such as
potential missing values or inconsistencies could also have affected the
accuracy of the findings here. Lastly, the linear regression models here
assume a linear relationship between predictors and donations which can
greatly oversimplify the complexities.

Future research should not only aim to address these limitations, but
expand the analysis to other provinces and territories across Canada.
This would enhance the generalisability of the results and provide a
more nuanced understanding of regional differences. Furthermore,
utilizing nonlinear modelling techniques or looking at a broader set of
predictor variables such as donor demographics or economic indicators
could better capture the nuances of donor behaviour and also mitigate
the risk of omitted variable bias. Comparative studies with other
democratic countries, beyond Canada, could provide more general
perspectives on how other political systems and regulatory frameworks
influence or are influenced by donation behaviours. Other valuable areas
of future exploration include assessing the impact of media coverage on
donation trends. Finally, by addressing these areas, research can build
on the current study's findings and contribute to a greater
understanding of political donations.

@@

**Idealized Methodology**

>>>>

A key question in political finance is whether a party's position in
power affects the amount of financial support it receives. Studies in
other countries, such as the United States, have shown that incumbents
often benefit from more donations, particularly from interest groups
seeking to maintain influence [@fouirnaies2014financial]. However, there
has been less focus on whether this holds true in Canada, where the
political system and donation regulations differ significantly. This
paper aims to fill that gap by analyzing how a party's power status
influences donor behaviour in federal and Ontario provincial elections.

@@

The analysis was conducted using both R [@citeR] and Python [@python],
leveraging a range of libraries for data manipulation and visualization.
In R, key packages included dplyr [@dplyr], lubridate [@lubridate],
tidyverse [@tidyverse], ggplot2 [@ggplot], readr [@readr], tibble
[@tibble], and rstanarm [@rstanarm] for statistical modeling. In Python,
pandas [@pandas], requests [@requests], and matplotlib [@matplotlib]
were utilized for gathering the data.

@@

At the heart of politics is finance, as revealed by Traag’s research on the complex contagion of campaign donations in the United States [@traag2016complex]. Traag’s study found that most campaign contributions come from a small number of wealthy donors linked by large social networks, demonstrating how a few individuals can shape political outcomes. Similar dynamics appear to be at play in Canada. Here, donors often prefer to support parties not currently in power. One reason might be the hope that challengers will bring needed policy changes. More realistically, as noted by Mollona [@mollona2022ethics], politicians, especially those out of power, often seek funding in exchange for promoting donors’ interests. Over time, this can lead to a pattern where donors shift their support to whoever best serves their goals, fostering continuous political competition.
A key finding of this study is that at both the Ontario and federal levels, the Conservative Party generally attracts more financial backing than other parties. This steady flow of resources may help the Conservatives maintain a strong presence in the political arena, influencing agenda-setting and public debate. With more funds, they can more easily finance voter outreach, shape their messaging, and solidify their support networks. While this financial advantage may help sustain dynamic election contests, it also underscores concerns about whose interests gain the most traction. When one party consistently out-raises others, the policy conversation may lean toward those who provide the most money, potentially sidelining issues important to groups with fewer financial resources.

## International Comparisons and the Importance of Regulation
These trends are not confined to Canada. In the United Kingdom, for example, large private donations, often from very wealthy “superdonors”, widen the gap between major parties and smaller competitors [@draca2022financing]. Such patterns highlight a broader democratic challenge, as money can concentrate power and influence political direction in ways that may not always reflect broad public preferences. Rather than serving purely as a resource that encourages healthy competition, political donations can also become a tool that benefits certain parties and voices more than others.
Regulation and transparency are essential. Clearer rules and better public reporting help citizens understand who funds whom and why. Drawing lessons from other countries may guide reforms aimed at limiting undue influence. Stronger oversight, spending caps, and easy-to-access donor databases can empower voters to make informed choices. Without these safeguards, the already powerful, whether individuals, industries, or interest groups, can gain even more sway, risking an erosion of trust in democratic institutions.

@@

This study provides valuable insights into donation behaviours across
Canada but it its limitations. First, the analysis is applied to only
federal and provincial donation records, overlooking regional variations
found in other provinces and territories. The geographical limitations
of this study hinder its ability to generalize across the entire
country. Moreover, the study only spans from 2006 to 2024 which economic
or political events could have influence over the donation patterns that
the models do not account for. Additionally qualitative factors such as
personal beliefs, specific policy preferences and campaign effectiveness
are not directly addressed in this study. Data quality issues such as
potential missing values or inconsistencies could also have affected the
accuracy of the findings here. Lastly, the linear regression models here
assume a linear relationship between predictors and donations which can oversimplify the complexities.

Future research should not only aim to address these limitations, but
expand the analysis to other provinces and territories across Canada.
This would enhance the generalizability of the results and provide a
more nuanced understanding of regional differences. Furthermore,
utilizing nonlinear modelling techniques or looking at a broader set of
predictor variables such as donor demographics or economic indicators
could better capture the nuances of donor behaviour and also mitigate
the risk of omitted variable bias. Comparative studies with other
democratic countries, beyond Canada, could provide more general
perspectives on how other political systems and regulatory frameworks
influence or are influenced by donation behaviours. Other valuable areas
of future exploration include assessing the impact of media coverage on
donation trends. Finally, by addressing these areas, research can build
on the current study's findings and contribute to a greater
understanding of political donations.

@@

##Idealized Methodology
